---
title: "Economic Predictors of House Incumbent Vote Share"
author: "Vivian Nguyen"
date: '2022-09-19'
output: pdf_document
categories: []
tags: []
slug: []
---

This is blog post #2 in a series of analytical posts in lieu of the 2022 midterms. This post is affiliated with Gov 1347: Election Analytics, a course at Harvard University in the department of Government.

---

```{r setup, include=FALSE}

## Set up

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(car)
library(readr)
library(stargazer)
library(gridExtra)

popvote_df <- read_csv('house_popvote_seats.csv') 
GDPquarterly_df <- read_csv('GDP_quarterly.csv')
RDIquarterly_df <- read_csv('RDI_quarterly.csv')
UEquarterly_df <- read_csv('unemployment_national_quarterly_final.csv')

```

```{r, include = FALSE, message=FALSE}

## Merge predictor and outcome data, filter for just the last Q of the election cycle

results_GDP <- left_join(GDPquarterly_df, popvote_df, by = 'year') %>% 
  drop_na() %>% 
    select('year', 'winner_party', 'H_incumbent_party', 'H_incumbent_party_majorvote_pct', 'quarter_cycle', 'GDP_growth_pct', 'GDP_growth_qt') %>%
    filter(quarter_cycle == 8)  

results_RDI <- left_join(RDIquarterly_df, popvote_df, by = 'year') %>% 
  drop_na() %>% 
    select('year', 'winner_party', 'H_incumbent_party', 'H_incumbent_party_majorvote_pct', 'quarter_cycle', 'DSPIC_change_pct', 'DSPIC_change_qt') %>%
    filter(quarter_cycle == 8) 

UEquarterly_df$UNRATE_change <- 0 

for (index in 1:nrow(UEquarterly_df)){
  if (index != 1)
    UEquarterly_df$UNRATE_change[index] <- UEquarterly_df$UNRATE[index] - UEquarterly_df$UNRATE[index - 1]
}

results_UE <- left_join(UEquarterly_df, popvote_df, by = 'year') %>% 
  drop_na()  %>% 
    select('year', 'winner_party', 'H_incumbent_party', 'H_incumbent_party_majorvote_pct', 'quarter_cycle', 'UNRATE', 
           'unrate_qt_total', 'UNRATE_change') %>%
    filter(quarter_cycle == 4) 

## Joining filtered datasets together

newdf <- left_join(results_UE, results_GDP, by = 'year')
newdf <-left_join (newdf, results_RDI, by = 'year') %>%
  select('year', 'winner_party.x', 'H_incumbent_party.x', 'H_incumbent_party_majorvote_pct.x', 'quarter_cycle.x', 'UNRATE', 'GDP_growth_pct', 'DSPIC_change_pct', 'unrate_qt_total', 'DSPIC_change_qt', 'GDP_growth_qt', 'UNRATE_change') %>%
  rename('winner_party' = 'winner_party.x', 'H_incumbent_party_majorvote_pct' = 'H_incumbent_party_majorvote_pct.x', 'quarter_cycle' = 'quarter_cycle.x', 'H_incumbent_party' = 'H_incumbent_party.x')

```

**Blog Extension: Model Evaluation. Build multiple predictive models using national economic variables as predictors. Compare those models using the tools we learned today. How much is your 2022 prediction sensitive to the change of measure(s)? What does it tell us about the economic model of voting behavior?**

Last week, my dive into the 2018 election results and swings confirmed that the president's party often performs poorly in the midterms. By the 2020 House elections, the president's party gained some of its vote share back. 

This week, I temporarily move away from cyclical election study to delve into one of the most promising ways to predict election outcomes -- economics. Retrospective voting, a model of voting in which voters elect leaders based off of recent experiences and judgment of welfare under incumbent leadership, dictates that for many elections, the results largely rely on the state of the economy in the last quarter of the election cycle (Achen & Bartels, 2017). For presidential elections, this means that national economic variables like GDP (gross domestic product), unemployment, and RDI (real disposable income) can account for a lot of the variation in incumbent candidate popularity. **I explore this relationship between national economic variables and incumbent vote share, but for House races from 1948 to 2020. I want to know which variables accurately predict House incumbent party vote share, and build a model accordingly. Lastly, I will test my model with data from 2020 to see how well it would've predicted the last election.** 

To start identifying which national economic variables predict House incumbent vote share well, I first investigated how Quarter 8 GDP growth (percentage), RDI change (percentage), and unemployment rate, individually, are related to incumbent party vote share. I specifically choose Quarter 8 because voters often refer to the latest memory they have of the economy to judge how the economy has performed under their incumbent leader(s) (Achen & Bartels, 2017. Healy & Lenz, 2014).

```{r, echo=FALSE, fig.width = 8, fig.height=6, out.width = '100%', fig.align='center'}

## Correlations

GDP_cor <- paste("corr(GDP, Y) = ", round(cor(results_GDP$GDP_growth_pct, results_GDP$H_incumbent_party_majorvote_pct), 4))
RDI_cor <- paste("corr(RDI, Y) = ", round(cor(results_RDI$DSPIC_change_pct, results_RDI$H_incumbent_party_majorvote_pct), 4))
UE_cor <- paste("corr(UE, Y) = ", round(cor(results_UE$UNRATE, results_UE$H_incumbent_party_majorvote_pct), 4))

## Simple plots of univariable regressions for GDP, RDI, and UE

# GDP plot
g1 <- newdf %>%
  ggplot(aes(x=GDP_growth_pct, y=H_incumbent_party_majorvote_pct,
             label=year)) + 
    geom_text() + geom_text(x=-1.2, y=44.7, label="Y = 52.42 - 0.9118 * X", color = 'red') + geom_text(x=2.3, y=44.7, label = GDP_cor, color = 'firebrick') +
    geom_smooth(method="lm", formula = y ~ x) +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.01, lty=2) + # median
    xlab("Q8-Q7 GDP growth percentage (X)") +
    ylab("Incumbent party PV (Y)") +
    theme_bw() + ylim(45, 58) +
    ggtitle("GDP Growth %") 

# RDI plot
g2 <- newdf %>%
  ggplot(aes(x = DSPIC_change_pct, y = H_incumbent_party_majorvote_pct,
             label=year)) + 
      geom_text() + geom_text(x=-1.1, y=44.7, label="Y = 52.58 - 0.7964 * X", color = 'red') + geom_text(x=2.7, y=44.7, label = RDI_cor, color = 'firebrick') +
    geom_smooth(method="lm", formula = y ~ x) +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.01, lty=2) + # median
    xlab("Q8-Q7 RDI change percentage (X)") +
    ylab("Incumbent party PV (Y)") +
    theme_bw() + ylim(45, 58) +
    ggtitle("RDI Change %")

# UE plot
g3 <- newdf %>%
  ggplot(aes(x = UNRATE, y = H_incumbent_party_majorvote_pct,
             label=year)) +
      geom_text() + geom_text(x=3.8, y=44.7, label="Y = 47.48 + 0.7466  * X", color = 'red') + geom_text(x=7, y=44.7, label = UE_cor, color = 'firebrick') +
    geom_smooth(method="lm", formula = y ~ x) +
    geom_hline(yintercept=50, lty=2) +
    geom_vline(xintercept=0.01, lty=2) + # median
    xlab("Q8-Q7 Unemployment rate (X)") +
    ylab("Incumbent party PV (Y)") +
    theme_bw() + xlim(2.7, 8) + ylim(45, 58) +
    ggtitle("Unemployment Rate")

grid.arrange(g1, g2, g3, ncol = 3)

```

As shown above, each variable, GDP growth, RDI change, and unemployment rate, respectively, are correlated to House incumbent vote share - as GDP growth percentage decreases, or RDI change percentage decreases, or unemployment rate grows, House incumbent vote share grows. These correlations are unintuitive, and suggest that voters do not accurately hold their leaders accountable according to national economic performance. 

That said, these correlations are not too explanatory by themselves and may be very susceptible to noise, with R-squared values of 0.081, 0.056, and 0.131. I moved on from univariate relationships between these national economic variables House incumbent vote share to multivariable relationships, hoping to find a model that incorporates multiple economic variables to predict incumbent vote share. In this multivariable model (which I call Model 1), I consider the predictors `GDP_growth_pct` (GDP growth %), `DSPIC_change_pct` (RDI change %), `GDP_growth_qt` (GDP growth), and `UNRATE` (unemployment rate) to investigate how they influence incumbent vote share in conjunction. 

```{r, echo=TRUE}

lm_GDP <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct, data = newdf)
lm_RDI <- lm(H_incumbent_party_majorvote_pct ~ DSPIC_change_pct, data = newdf)
lm_UE <- lm(H_incumbent_party_majorvote_pct ~ UNRATE, data = newdf)
lm_model <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct + DSPIC_change_pct +  GDP_growth_qt + UNRATE, data = newdf)

stargazer(lm_GDP, lm_RDI, lm_UE, lm_model, title = "Results", type = "text", column.labels = c("GDP", "RDI", "UE", "Model 1"))

```
```{r, include = FALSE}

summary(lm_GDP)$r.squared
summary(lm_RDI)$r.squared
summary(lm_UE)$r.squared
summary(lm_model)$r.squared

```
My development of Model 1 finds that GDP, RDI, and unemployment all influence House incumbent vote share, but in varying directions and with varying levels of statistical significance. For GDP growth (%), Model 1 anticipates that as it grows by 1, House incumbent vote share will too by 2.9, which makes more sense. However, for GDP growth (absolute) and unemployment rate, increases in these are still associated with decreases in incumbent vote share. 

These results are again fascinating, as they are counter-intuitive and don't follow the traditional knowledge political scientists lean on concerning retrospective voting, economic performance, and incumbent success. That said, Model 1 has an R-squared value of 0.386, which is much higher than the univariate R-squared values previously found. 

```{r, fig.width=3, fig.height=2, out.width = '30%', fig.align='center'}

# My model
plot(newdf$year, newdf$H_incumbent_party_majorvote_pct, 
     type="l",
     main="Model 1 Predictions",
     sub = "True Y (line), Predicted Y (dot) for each year",
     col.sub = "firebrick",
     xlab = "Year",
     ylab = "House Incumbent Party Vote Share")
points(newdf$year, predict(lm_model, newdf), col = "red", pch = 18)

```

To evaluate the performance of my model and choice of model, I check the residuals. As shown below, they are roughly normally distributed around a center point of 0. This tells me that the form of Model 1 (linear) is adequate.

```{r, fig.width=6, fig.height=4, out.width = '70%', fig.align='center'}
# Residual distribution of my model
hist(lm_model$model$H_incumbent_party_majorvote_pct - lm_model$fitted.values,
     main="Histogram of Residuals (true Y - predicted Y)", xlab = "Residual")
```

```{r}
outsamp_errors <- sapply(1:1000, function(i){
    years_outsamp <- sample(newdf$year, 8)
  outsamp_mod <- lm(H_incumbent_party_majorvote_pct ~
                      GDP_growth_pct,
                  newdf[!(newdf$year %in% years_outsamp),])
  outsamp_pred <- predict(outsamp_mod,
                newdata = newdf[newdf$year %in% years_outsamp,])
  outsamp_true <- newdf$H_incumbent_party_majorvote_pct[newdf$year
                                                       %in% years_outsamp]
  mean(outsamp_pred - outsamp_true)
})
```

```{r, echo=FALSE, message=FALSE, fig.width=6, fig.height=3, out.width = '70%', fig.align='center'}
hist(outsamp_errors,
     xlab = "",
     main = "Mean Out-of-Sample Residual\n(1000 runs of cross-validation)")
```
After checking that Model 1 performs well after being tested with 1,000 runs of cross-validation, with a mean-error of roughly 1, I move to applying Model 1 to the most recent Q8 data available for the national economic variables in the model. This data comes from 2020, where


```{r, echo=TRUE, eval=FALSE}

GDP_new <- GDPquarterly_df %>%
    subset(year == 2020 & quarter_cycle == 8) %>%
    select(GDP_growth_pct)
UNRATE_new <- UEquarterly_df %>%
    subset(year == 2020 & quarter_cycle == 4) %>%
    select(UNRATE)
DSPIC_new <- RDIquarterly_df %>%
    subset(year == 2020 & quarter_cycle == 8) %>%
    select(DSPIC_change_pct)
GDP_qt_new <- GDPquarterly_df %>%
    subset(year == 2020 & quarter_cycle == 8) %>%
    select(GDP_growth_qt)

new_data <- data.frame(GDP_new, DSPIC_new, GDP_qt_new, UNRATE_new)

predict(lm_model, new_data, interval="confidence")

newdf %>%
  filter(year == 2020) %>%
  summarise(actual = H_incumbent_party_majorvote_pct)

#Interpretation: This means that, according to our model, a year with these traits, on average, has a (stopping distance ranging between 51.83 and 62.44 ft)46, 54) H incumbent vote share

```









Though my model predicted the House popular vote share of 2020 well, even given the novel nature of the economics of that COVID-ridden election cycle, it still can be improved with wider purview into House incumbent seat share and totals. As discussion in Gov 1347: Election Analytics, for House races, it is often the actual seats in the House that matter to parties and party members. Due 

Wright wrote in 2012 that worsened (so, increased) unemployment affects Democrats differently than it does Republicans. I wonder if this finding holds when we focus specifically on House elections -- this would be interesting to explore given more time.



---

**References**

[1] Achen, & Bartels, L. M. (2017). Democracy for realists (REV - Revised). Princeton University Press.
https://muse.jhu.edu/book/64646

[2] Healy, & Lenz, G. S. (2014). Substituting the End for the Whole: Why Voters Respond Primarily to the Election-Year Economy. American Journal of Political Science, 58(1), 31–47. https://doi.org/10.1111/ajps.12053

[3] Wright. (2012). Unemployment and the Democratic Electoral Advantage. The American Political Science Review, 106(4), 685–702. https://doi.org/10.1017/S0003055412000330

[4] Rakich, N., & Mejia, E. (2022, March 31). The house map's Republican bias will plummet in 2022 - because of gerrymandering. FiveThirtyEight.  https://fivethirtyeight.com/features/the-house-maps-republican-bias-will-plummet-in-2022-because-of-gerrymandering/ 
