---
title: The Ground Game
author: Vivian Nguyen
date: '2022-10-17'
slug: []
categories: []
tags: []
---

This is blog post #6 in a series of analytical posts in lieu of the 2022 midterms. This post is affiliated with Gov 1347: Election Analytics, a course at Harvard University in the department of Government.

---
```{r setup, include = FALSE, message = FALSE}

# Set up
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(readr)
library(stargazer)
library(usmap)
library(rmapshaper)
library(sf)
library(insight)
library(scales)

# Load in expert ratings and district polls
incumb <- read_csv("incumb_dist_1948-2020.csv") %>%
  janitor::clean_names()
house_polls <- read_csv("house_polls_long.csv") %>%
  janitor::clean_names()
house_cands <- read_csv("house_cands.csv") %>%
  janitor::clean_names()
house_cands_22 <- read_csv("house_cands_2022.csv") %>%
  janitor::clean_names()
cvap <- read_csv("cvap_district_2012-2020.csv") %>%
  janitor::clean_names()
df <- read_csv("turnout_2012_2020.csv")

```
# The Plan This Week
Last week, I investigated the impact of television advertisements on party success at the district level, and found a small, arguably negligible, effect. This week, I delve into the other major side of campaign activities – “the ground game.” Do the phone banks, text banks, rallies, door knockers, and more matter? I will incorporate these findings and retroactive ones (expert predictions from [week 4](https://vivian-1372.github.io/Election-Analytics/post/2022-10-03-incumbency)) to make updates to my district-level model, newly featuring the effect of the ground game, expert predictions, and candidate incumbency. This will produce new district-level predictions for the races finalizing in the next 3 weeks!    

# The Ground Game
As mentioned in my discussion of campaigns last week, political campaigns aim to mobilize supporters to vote, persuade potential voters to cast their ballot in favor of certain candidates, and encourage donations. In a 2018 study, it was found that across 49 field experiments, “the best estimate of the effects of campaign contact and advertising on Americans’ candidates choices in general elections is zero” (Kalla and Broockman). At the end of this research paper, Kalla and Broockman assert that though the ground game doesn’t seem to persuade voters to change their vote choice, these ground campaigns matter in other ways. 

The consensus from the literature and pundit commentary is that the ground game – the fundraising, voter mobilization, person-to-person contact, field offices, and bottom-up political strategy of campaigns – does indeed matter. It certainly did for Obama, whose campaigns were studied extensively by political scientists. In 2012, turnout increased by about 7 percentage points in states that were highly-targeted by the ground arm of the major campaigns (Enos and Fowler, 2018). Even as early as 2008, President Obama’s campaigns were notably strong. [The Atlantic](https://www.theatlantic.com/politics/archive/2012/10/obamas-edge-the-ground-game-that-could-put-him-over-the-top/264031/) praised Obama’s “sophisticated organization,” which debuted in 2008 as “the largest grassroots organization in the history of American politics” (Ball, 2012). A 2009 study found that counties with Obama campaign field offices had disproportionate increases in Democratic vote share (Masket), and a 2014 study found that the field offices increased county-level vote share by 1%, which in turn scored Obama ~275,000 votes in the 2008 election (Darr and Levendusky). 

These findings suggest that successful ground campaigns increase turnout. In the remainder of my blog post this week, I want to explore the question of how increased turnout at the district level affects the parties’ respective vote shares, if at all? Does higher Democrat turnout correlate with better Democrat major vote share – which is rather intuitive –, or no – maybe higher Democrat turnout occurs with more Republican turnout, so there’s a “cancel-out” effect?  

# My 2022 Model and Forecast, Updated
```{r 435 models, include = FALSE}

models = unique(house_cands_22$st_cd_fips)

for (stcdfip in models){
  model = lm()
  models <- c(models, model)
  print(stcdfip)
}

```
My previous district-level model was basic, relying on only average expert predictions. Though the predictions themselves consider many important electoral predictors, their aggregation and then averaging render them basic. This week, I aim to Incorporate turnout, incumbency and expert predictions into my district-level model of two-party major vote share.

## My New Model
Reflect on new models 

## My New Forecast
Present my predictions for the districts this year using 2022 data. 

---
**References**

[1] Kalla, & Broockman, D. E. (2018). The Minimal Persuasive Effects of Campaign Contact in General Elections: Evidence from 49 Field Experiments. The American Political Science Review, 112(1), 148–166. https://doi.org/10.1017/S0003055417000363

[2] Ball, M. (2012, October 25). Obama's edge: The ground game that could put him over the top. The Atlantic. https://www.theatlantic.com/politics/archive/2012/10/obamas-edge-the-ground-game-that-could-put-him-over-the-top/264031/) 

[3] Enos, & Fowler, A. (2018). Aggregate Effects of Large-Scale Campaigns on Voter Turnout. Political Science Research and Methods, 6(4), 733–751. https://doi.org/10.1017/psrm.2016.21

[4] Masket, Seth E. (2009). Did Obama’s Ground Game Matter? The Influence of Local Field Offices During the 2008 Presidential Election, Public Opinion Quarterly, 73(5), 1023–1039. https://doi.org/10.1093/poq/nfp077

[5] Darr, & Levendusky, M. S. (2014). Relying on the Ground Game. American Politics Research, 42(3), 529–548. https://doi.org/10.1177/1532673X13500520


